---
title: "Introduction to R"
subtitle: "R Markdown demo"
author: 
- "Clémentine Cottineau"
- "Claudiu Forgaci"
date: "`r format(Sys.time(), '%d %B %Y')`" # to be replaced with the date of the session
output:
    html_document:
        keep_md: yes
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
	echo = TRUE,
	message = FALSE,
	warning = FALSE
)
```

## Introduction {#intro}
This document is meant to demonstrate the use of R in an R Markdown document. We show a basic process of data collection, visualisation of non-spatial and spatial data, and interpretation.

We start by loading the packages that will be used in this notebook.

```{r load-packages, message=FALSE}
library(tidyverse)  # collection of packages for data manipulation, visualisation, etc.
library(googlesheets4)  # access data from Google sheets
library(sf)  # working with spatial data
library(osmdata)  # access OpenStreetMap data via the Overpass API
library(osmextract)  # retrieve bulk OpenStreetMap data
library(ggmap)
library(kableExtra)  # create and style HTML tables
```

## Reading the data {#read-data}

*NOTE: Clementine, here I included two ways to read in the data. I suggest we go for the second option. If you agree, we can remove this part from the Rmd.*

One way to get the survey data would be to read it in directly from `csv` files. When the goal is to collect responses through a survey, this approach has a few disadvantages:
- respondents need to fill in the sheets correctly, i.e., without using the separator (e.g. a comma) used by the `csv` file; 
- `csv` files are not 'friendly', that is, the data is shown in plain text, with fields separated by ',' or ';', not in tables;
- files are either named differently, or respondents need to name the files after 
- each respondent creates a separate file, which need to be read in and combined for analysis;
- after filling in the survey, respondents need to send the `csv` files by email.

```{r read-survey-data-csv}
# read survey data from csv files
files <- list.files(pattern="^survey-")
survey <- tibble()
for (i in seq_along(files)){
    survey <- rbind(survey, t(read_csv(files[i])[,2]))
}
```

This approach is error-prone and involves too many steps. Luckily, the R ecosystem of packages offers several other solutions to import survey data. As Google forms allows us to easily deploy a survey and to store the data in Google sheets^[], we will use the `googlesheets4` package to read in the survey data from there. Filling in the survey is user-friendly and the obtained data is in the right format: a so-called **data frame** with entries (survey responses) stored in rows and variables (survey questions) stored in columns.

```{r read-survey-data-google}
# read survey data from a Google sheet that stores the survey responses
# note that getting data via `googlesheets4` requires a Google account
survey_google <- read_sheet("https://docs.google.com/spreadsheets/d/1_eErh6dy7uMiEewDc4OW_R3sl0qah2-1kYj3JW9H5tA/edit?resourcekey#gid=800252241")

# clean survey data
colnames(survey_google) <- c("timestamp", "name", "year_start_tud", "place", "interest_r", "progr_exp")
survey_google <- survey_google %>%
    mutate(progr_exp = ifelse(progr_exp == "Yes", TRUE, FALSE)) 

# print cleaned survey data
survey_google
survey_google %>% 
    kbl() %>%
    kable_styling(bootstrap_options = c("striped", "hover"))
```

## Making graphs {#graphs}

## Making maps {#maps}

*NOTE: This part still needs work: (1) remove the script to a separate `plot_places.R` file and source it here; otherwise the script will be too long; (2) the `osmdata` script is limited by the download rate of the Overpass API; consider caching the results and if that does not work try to switch to the `osmextract` package in the next code chunk.*

The following code takes your home towns from the online survey, processes them in a few geoprocessing steps (like you would do in QGIS) and adds them as figures in the R Markdown document. The code is sourced from a separate `render_places.R` script which does the following:

1. it retrieves a set of layers from OpenStreetMap for the place you named in your survey
1. it clips the layers to the boundaries of the municipality
1. it styles the map

<!-- Parts of the code were adapted from: -->

<!-- - http://joshuamccrain.com/tutorials/maps/streets_tutorial.html -->
<!-- - http://estebanmoro.org/post/2020-10-19-personal-art-map-with-r/ -->
<!-- - https://ggplot2tutor.com/tutorials/streetmaps -->


```{r places-osm-1, include=FALSE}
places <- survey[, 2]
places_google <- survey_google$place

# source("render_places.R")

# 1. Get city boundaries
get_boundaries <- function(places) {
    for (i in seq_along(places)) {
        bb <- getbb(
            place_name = places[i],
            format_out = 'sf_polygon',
            silent = FALSE
        )
        
        big_streets <- getbb(places[i]) %>%
            opq() %>%
            add_osm_feature(
                key = "highway",
                value = c(
                    "motorway",
                    "primary",
                    "motorway_link",
                    "primary_link"
                )
            ) %>%
            osmdata_sf() %>%
            pluck("osm_lines") %>%
            st_intersection(bb)
        
        med_streets <- getbb(places[i]) %>%
            opq() %>%
            add_osm_feature(
                key = "highway",
                value = c(
                    "secondary",
                    "tertiary",
                    "secondary_link",
                    "tertiary_link"
                )
            ) %>%
            osmdata_sf() %>%
            pluck("osm_lines") %>%
            st_intersection(bb)
        
        small_streets <- getbb(places[i]) %>%
            opq() %>%
            add_osm_feature(
                key = "highway",
                value = c(
                    "residential",
                    "living_street",
                    "unclassified",
                    "service",
                    "footway"
                )
            ) %>%
            osmdata_sf() %>%
            pluck("osm_lines") %>%
            st_intersection(bb)
        
        river <- getbb(places[i]) %>%
            opq() %>%
            add_osm_feature(key = "natural",
                            value = "water") %>%
            # add_osm_feature(key = "water",
            #                 value = "river") %>%
            osmdata_sf() %>%
            pluck("osm_polygons") %>%
            st_intersection(bb)
        
        railway <- getbb(places[i]) %>%
            opq() %>%
            add_osm_feature(key = "railway", value = "rail") %>%
            osmdata_sf() %>%
            pluck("osm_lines") %>%
            st_intersection(bb)
        
        ggplot() +
            geom_sf(
                data = river,
                inherit.aes = FALSE,
                fill = "steelblue",
                color = "white",
                size = 0.05,
            ) +
            geom_sf(
                data = railway,
                inherit.aes = FALSE,
                color = "orange",
                size = .6,
                # linetype = "dotdash",
            ) +
            geom_sf(
                data = med_streets,
                inherit.aes = FALSE,
                color = "black",
                size = .3,
            ) +
            geom_sf(
                data = small_streets,
                inherit.aes = FALSE,
                color = "#666666",
                size = .2,
            ) +
            geom_sf(
                data = big_streets,
                inherit.aes = FALSE,
                color = "black",
                size = .8,
            ) +
            geom_sf(
                data = st_cast(bb, "LINESTRING"),
                inherit.aes = FALSE,
                color = "darkred",
                size = 1
            ) +
            coord_sf(
                xlim = getbb(places[i])[1, ],
                ylim = getbb(places[i])[2, ],
                expand = FALSE
            ) +
            theme_void() +
            # theme(
            #     plot.title = element_text(
            #         size = 20,
            #         family = "lato",
            #         face = "bold",
            #         hjust = .5
            #     ),
            #     plot.subtitle = element_text(
            #         family = "lato",
            #         size = 8,
            #         hjust = .5,
            #         margin = margin(2, 0, 5, 0)
            #     )
            # ) +
            labs(title = toupper(places[i]))
    }
}

get_boundaries(places[1]) # only run this for Delft; otherwise it exceeds the Overpass API rate limit
```
Alternative code with the `osmextract` package.

```{r places-osm-2, echo=TRUE}
library(sf)
library(osmextract)
library(osmdata)

# get places from csv survey
places <- survey[, 2]
# get places from google survey
places_google <- survey_google$place

# create map - test Delft
## get municipal boundary from osmdata
bb2 <- osmdata::getbb(
            place_name = "Delft, Netherlands",
            format_out = 'sf_polygon',
            silent = FALSE
        )

## get data from osmextract
osm_lines = oe_get("Delft, Netherlands", stringsAsFactors = FALSE, quiet = TRUE)
nrow(osm_lines)
ht <- c("primary", "secondary", "tertiary", "unclassified") # highway types of interest
osm_major_roads_delft <- osm_lines[osm_lines$highway %in% ht, ] %>% 
    sf::st_intersection(bb2)

## plot data
plot(osm_major_roads_delft["highway"], key.pos = 1)
```

