---
title: "Introduction to R"
subtitle: "R Markdown demo"
author: 
- "Clémentine Cottineau"
- "Claudiu Forgaci"
date: "`r format(Sys.time(), '%d %B %Y')`" # to be replaced with the date of the session
output:
    html_document:
        keep_md: yes
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
	echo = TRUE,
	message = FALSE,
	warning = FALSE
)
```

## Introduction {#intro}
This document is meant to demonstrate the practical use of R for automation, statistics, visualisation and reproducibility in an *R Markdown document*. We show basic examples of data collection through a computation loop, visualisation of non-spatial and spatial data, and interpretation.

We start by loading the packages that will be used in this notebook.

```{r load-packages, message=FALSE}
library(tidyverse)  # collection of packages for data manipulation, visualisation, etc.
library(googlesheets4)  # access data from Google sheets
library(sf)  # working with spatial data
library(osmdata)  # access OpenStreetMap data via the Overpass API
library(osmextract)  # retrieve bulk OpenStreetMap data
library(ggmap)
library(kableExtra)  # create and style HTML tables
```

## Reading the data {#read-data}

One way to get the survey data would be to read it from several `csv` files, for instance information received from students in a course, or from participants in a research project.

```{r read-survey-data-csv}
# lists the files with "survey-" in their name in the same folder as this document
files <- list.files(pattern="^survey-")  
files
```

For this, we will use a 'for-loop', i.e. a simple recursive function very useful to automate operations.

```{r for-loop-data-csv}
# creates an empty table
survey <- tibble() 

n <- length(files)
# the 'for-loop':
for (i in 1:n){
  f <- read.csv2(files[i], header=F)
  if(i == 1){
    survey <- f
  } else {
    survey <- cbind(survey, f[,2])
  }
}

transpose <- as.data.frame(t(survey))
colnames(transpose) <- c("Name", "ArrivalYearTU", "CityBefore", "InterestR", "Experience")
survey_together <- transpose[-1,]
rownames(survey_together) <- 1:n
 survey_together$ArrivalYearTU <- as.numeric(survey_together$ArrivalYearTU)
 survey_together$InterestR <- as.numeric(survey_together$InterestR)
 
 # as with spreadsheets, we can create new variables
 survey_together$SeniorityAtTUD <- 2021 - survey_together$ArrivalYearTU
survey_together 
```

It still goes quicker to do it by hand for 3 files, but when the number of files increases (or when files change), the code is the same, and this is when it makes sense to automate the process.

Moreover, when the goal is to collect responses through a survey, working with `csv` files has a few disadvantages:
- respondents need to fill in the sheets correctly, i.e., without using the separator (e.g. a comma) used by the `csv` file, or there is more work to clean the data;
- `csv` files are not 'friendly', that is, unless you "open with" a spreadsheet viewer, the data is shown in plain text, with fields separated by ',' or ';', not in tables;
- files are either named differently, or respondents need to name the files after 
- each respondent creates a separate file, which need to be read in and combined for analysis;
- after filling in the survey, respondents need to send the `csv` files by email.

This approach is error-prone and involves too many steps. Luckily, the R ecosystem of packages offers several other solutions to import survey data. As Google forms allows us to easily deploy a survey and to store the data in Google sheets, we will use the `googlesheets4` package to read in the survey data from there. Filling in the survey is user-friendly and the obtained data is in the right format: a so-called **data frame** with entries (survey responses) stored in rows and variables (survey questions) stored in columns.

**> Let's try with your surveys!**

```{r read-survey-data-google}
# Read survey data from a Google sheet that stores the survey responses
# Note that getting data via `googlesheets4` requires a Google account
# unless `gs4_deauth()` is used. Here we choose to deauthorise so that
# the workshop participants can easily access the data.
gs4_deauth()

survey_google <- read_sheet("https://docs.google.com/spreadsheets/d/1_eErh6dy7uMiEewDc4OW_R3sl0qah2-1kYj3JW9H5tA/edit?resourcekey#gid=800252241")

# clean survey data
colnames(survey_google) <- c("timestamp", "name", "year_start_tud", "place", "interest_r", "progr_exp")
survey_google <- survey_google %>%
    mutate(progr_exp = ifelse(progr_exp == "Yes", TRUE, FALSE)) %>% 
    filter(!grepl("Test", name))

# print cleaned survey data
survey_google
survey_google %>% 
    kbl() %>%
    kable_styling(bootstrap_options = c("striped", "hover"))
```

## Visualising non-spatial data: Making graphs {#graphs}

Let's visualise some non spatial data, for instance the relation between the semiority at TU Delft and the reported interest in R.

```{r graphs}
library(ggplot2)

q <- ggplot(survey_together, aes(x=SeniorityAtTUD, y=InterestR))
q + geom_point()
```

We can even estimate a linear regression between the two and visualise the result.

```{r regression}

fit <- lm(InterestR ~ SeniorityAtTUD, data = survey_together)

ggplot(fit$model, aes_string(x = names(fit$model)[2], y = names(fit$model)[1])) + 
  geom_point() +
  stat_smooth(method = "lm", col = "red") +
  labs(title = paste("Adj R2 = ",signif(summary(fit)$adj.r.squared, 2),
                     "Intercept =",signif(fit$coef[[1]],2 ),
                     " Slope =",signif(fit$coef[[2]], 2),
                     " P =",signif(summary(fit)$coef[2,4], 2)))

```

## Visualising spatial data: Making maps {#maps}

The following code takes your home towns from the online survey and displays them on an interactive map.

```{r places-osm-2, echo=TRUE}
library(sf)
library(osmextract)
library(osmdata)
library(leaflet)

# get places from csv survey
places <- survey_google[, 4]

for (i in 1:nrow(places)){
## get municipal boundary from osmdata
bb <- osmdata::getbb(
            place_name = places[i,],
            format_out = 'sf_polygon',
            silent = FALSE
        )
  assign(paste0("bb", i), bb)
  if(i == 1){
    bbx <- bb
  } else{
    bbx <- rbind(bbx, bb)
  }
}

#map places on interactive map
leaflet(bbx) %>% addProviderTiles("CartoDB.Positron") %>%
    addPolygons()
```

Let's extract the boundary and primary streets of the city of Rotterdam.
    
```{r places-osm-3}
## create a function to extract the streets of a city x
map_city_streets <- function(x) {
    osm_lines = oe_get(x, stringsAsFactors = FALSE, quiet = TRUE)
    nrow(osm_lines)
    # get highway types of interest
    # ht <- c("primary", "secondary", "tertiary", "unclassified") 
    ht <- c("motorway", "motorway_link", "primary", "primary_link") 
    bb <- osmdata::getbb(place_name = x,
                         format_out = 'sf_polygon',
                         silent = FALSE)
    osm_major_roads_place <- osm_lines[osm_lines$highway %in% ht,] %>%
        sf::st_intersection(bb)
    
    plot(bb[[1]])
    plot(osm_major_roads_place$geometry, add = T, col = "red")
}
    
## let's map it for Rotterdam
map_city_streets("Rotterdam, NL")
```

Let's use the same function on the last survey response.
 
```{r places-osm-4}
sl <- nrow(survey_google)
# map_city_streets(survey_google[, 4][[sl-2:sl]])
map_city_streets(survey_google[, 4][[sl]])
```

For larger OSM datasets, the `osmdata` pacakge might not be useful as it has a request limit. An alternative is the `osmextract` package which downloads entire OSM extracts to a local folder and retreives the spatial data from there. Let's see an example with the city of Delft.

```{r places-osm-5, echo=TRUE}
library(sf)
library(osmextract)
library(osmdata)

# get places from csv survey
# places <- survey[, 2]
# get places from google survey
# places_google <- pull(survey_google[, 4])

map_city_streets2 <- function(x) {
    ## get municipal boundary from osmdata
    bb <- osmdata::getbb(place_name = x,
                         format_out = 'sf_polygon',
                         silent = FALSE)
    
    ## get data from osmextract
    osm_lines = oe_get(x,
                       stringsAsFactors = FALSE,
                       # force_download = TRUE,
                       quiet = FALSE)
    nrow(osm_lines)
    ## get highway types of interest
    ht <- c(
        "motorway",
        "motorway_link",
        "primary",
        "primary_link",
        "secondary",
        "tertiary",
        "unclassified"
    )
    osm_major_roads <- osm_lines[osm_lines$highway %in% ht, ] %>%
        sf::st_intersection(bb)
    
    ## plot data
    plot(bb)
    plot(
        osm_major_roads["highway"]$geometry,
        key.pos = 1,
        add = T,
        col = "red",
        lwd = 0.7
    )
}

# map Delft
map_city_streets2("Delft, Netherlands")
```

Now let's use the same function on the last entry of the survey data.

```{r places-osm-6, echo=TRUE}
sl <- nrow(survey_google)
# map_city_streets(survey_google[, 4][[sl-2:sl]])
map_city_streets2(survey_google[, 4][[sl]])
```

To summarise, the code chunks in this section took your home towns from the online survey, processed them in a few geoprocessing steps (like you would do in QGIS) and added them as figures or interactive maps in the R Markdown document. General steps taken:

1. retrieve a set of layers from OpenStreetMap for the place you named in your survey
1. clip the layers to the boundaries of the municipality
1. style and display the map

<!-- Parts of the code were adapted from: -->

<!-- - http://joshuamccrain.com/tutorials/maps/streets_tutorial.html -->
<!-- - http://estebanmoro.org/post/2020-10-19-personal-art-map-with-r/ -->
<!-- - https://ggplot2tutor.com/tutorials/streetmaps -->
